Variation in gaze following across the life span: A process-level perspective

### 2024-03-27 paper/manuscript.Rmd
prepare revision

---
title: "Variation in gaze understanding across the life span: A process-level perspective"
shorttitle: "modeling variation in gaze understanding"
author:

  - name          : "Julia Christin Prein"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Max Planck Institute for Evolutionary Anthropology, Deutscher Platz 6, 04103 Leipzig, Germany"
    email         : "julia_prein@eva.mpg.de"
    role          :
      - Conceptualization
      - Methodology
      - Software
      - Investigation
      - Formal Analysis
      - Writing - Original Draft Preparation
      - Writing - Review & Editing

@@

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze-following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze understanding across the lifespan (*N* = 471 3- to 80-years-olds; children from a mid-sized German city; international remotely tested adults). We found a steep learning curve during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze understanding in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we tested two implications of our gaze model, namely that gaze understanding could be related to vector estimation in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze understanding is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "7914"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

# Variation in gaze understanding across the life span: A process-level perspective

@@

**Data availability statement:** The web application (<https://ccp-odc.eva.mpg.de/tango-demo/>) described here is open source (<https://github.com/ccp-eva/tango-demo>). The data sets generated during and/or analyzed during the current study are available in the [gazecues-modeling] repository (<https://github.com/ccp-eva/gazecues-modeling>). All experiments were pre-registered (<https://osf.io/zjhsc/>).

@@

**Funding:** This study was funded by the Max Planck Society for the Advancement of Science, a noncommercial, publicly financed scientific organization (no grant number). Manuel Bohn was supported by a Jacobs Foundation Research Fellowship.

@@

-	Gaze understanding develops beyond infancy. Highest precision levels in localizing attentional foci are reached in young adulthood with a slight decrease towards old age. 
-	We present a computational model as a mechanistic description of gaze understanding. The model explains individual differences and recovers signature patterns in the data.
-	Our model assumes that gaze understanding relies on vector estimation. To test this assumption experimentally, we designed a non-social vector estimation task. 
-	We found substantial correlations between gaze understanding and vector estimation, as well as Level 2 perspective-taking. Other Theory of Mind tasks did not correlate.

# Introduction
Following the gaze of others is valuable to extract information from the environment. It guides us to "informational hotspots" [@meltzoff2010social, p.1] and can be used to identify internal states such as intentions or emotions [@corkum1998origins; @pfeiffer2013gaze]. As one of the most fundamental social-cognitive abilities, gaze-following is an integral part of almost every form of social interaction, including communication, collaboration, cultural and social learning [@emery2000eyes; @rakoczy2022foundations; @frith2012mechanisms; @tomasello2003makes; @bohn2019pervasive; @moore2008development; @hessels2020how], and has been extensively studied in infancy [for review, see @delbianco2019developmental]. In the most commonly used paradigm [e.g., @astor2020social; @gredeback2010development; @byers-heinlein2021development; @ishikawaphysiological], the experimenter looks directly at the infant before shifting their head and eyes to one of the two objects. Infants’ looking times to the target or the proportion of choosing the target over the distractor are measured. Analyses traditionally focus on the average age at which children as a group reach an above-chance performance. Research in this tradition finds that infants as young as six months can follow gaze to that of another agent [@dentremont1997demonstration]. At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access [@moll200412; @deak2000effects; @butterworth1991minds; @corkum1995development].

While the emergence of gaze-following has been well established, less is known about its developmental trajectory beyond infancy. One possibility is that the ability to follow gaze does not improve once it emerges. Yet, most, if not all, cognitive abilities continue to develop throughout childhood [e.g., @gathercole2004structure; @gredeback2010development]. It seems likely that children fine-tune their gaze understanding as they get older - presumably while using them in social interactions. To capture development in gaze understanding beyond infancy, we conducted our Study 1 across the lifespan.

Up until today, only a handful of gaze-following studies differentiate between manipulating head and eye movement. @michel2021effects found that gaze-following in four-month-olds was likely driven by other's head- instead of eye movements. @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. From farther distance, body or face orientation can act as more salient cues to determine another's area of attention. However, eye direction indicates a more precise location of focus [@emery2000eyes], and allows to anticipate likely future actions [@zohary2022gaze; @friesen2011gaze]. In our three studies, we therefore focused on subtle gaze cues and isolated eye movement alone.

Though previous studies suggest that young infants can align their visual attention to another’s line of sight, this does not necessarily indicate understanding the intentions of the agent [@aslin2007look]. Infants could simply align their orientation without processing what exactly the other is seeing [@butterworth1991minds]. In fact, one might question if such an alignment reflects an understanding of visual perspectives at all, because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. To fulfill this criterion, @moll2006level applied an active behavioral choice, in which participants searched for a visually accessible or inaccessible target object. 24-month-olds succeeded in the task, while 18-month-olds did not. In an object choice study with two containers, @behne2005oneyearolds found that already 14-month-olds used an experimenter's gaze; however, this was again confounded with a congruent face movement [cf. @povinelli1997exploitation]. In our three studies, we assessed how subtle gaze cues guided participants' behavior in actively locating a target.

Finally, group-level analyses of gaze-following abilities may mask individual differences between children. Measuring individual differences in basic aspects of social cognition is especially important to understand the underlying processes and to quantify the impact of environmental influences and other cognitive abilities [@birch2017perspectives; @delbianco2019developmental]. In our studies, we measured gaze understanding continuously by using a task which is designed to capture individual-level variation [@prein2023tango].

The present study had two main goals: first, we studied the development of gaze understanding beyond infancy (Study 1). Instead of capturing the youngest age at which children follow gaze, we examined how this ability changes with age. Our second goal was to provide a process-level theory of gaze understanding (and individual differences therein) as a computational cognitive model. We formalized gaze understanding as a form of vector estimation and tested whether our model explained empirical data (Study 2). Finally, we examined two empirical predictions that followed from the model, namely that gaze understanding is related to non-social vector estimation and visual perspective-taking (Study 3).

# Study 1: Gaze understanding across the lifespan

The study was pre-registered at: <https://osf.io/snju6> (child sample) and <https://osf.io/6yjz3> (adult sample). The study obtained ethical clearance by the MPG Ethics commission Munich, Germany, falling under an umbrella ethics application (Appl. No. 2021_45). Data was collected between May 2021 and April 2023.

@@

We collected data online from 3- to 80- year-olds (see Supplements for further details). The child sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants and was recruited via an internal database of families in Leipzig, Germany, who volunteered to participate in child development studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (∼600,000 inhabitants; median individual monthly net income approx. 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

@@

We used the continuous version of the TANGO [@prein2023tango]. The task was presented as a web application (demo [https://ccp-odc.eva.mpg.de/tango-demo/](https://ccp-odc.eva.mpg.de/tango-demo/.); source code <https://github.com/ccp-eva/tango-demo>). The TANGO showed satisfactory internal consistency and retest reliability for children and adults (reliability estimates *Pearson's r* from .7 to .8; [@prein2023tango]).

@@

We fit GLMMs that make different assumptions about the developmental trajectory, modeling the relationship between age and performance as linear, quadratic, or cubic. In addition, we fit a Gaussian Process model [@burkner2017brms], which assumes a smooth relationship but avoids enforcing a particular shape. Per individuum, imprecision was aggregated across trials and modeled as a `lognormal` distribution.[^1] The unit of imprecision was counted in target widths, i.e., an imprecision of 1 meant clicking one balloon width to the left or right of the true target center. We inspected the posterior distributions (mean and 95% Credible Interval (CrI)) for the age estimates and compared models via model weights and the difference in expected log pointwise predictive density (ELPD) estimated using leave-one-out cross-validation (LOO) [@vehtari2017practicala].

[^1]: Originally, we fit our models on a trial-by-trial basis with the following structure: `performance ~ age + symmetricPosition + trialNr + (1 + symmetricPosition + trialNr | subjID)`. However, the Gaussian Process model was computationally heavy. Thus, we simplified the model structure, aggregated data on a subject level, and included only age as an effect. See Supplements for a comparison between the original and the here-reported model structures. The model predictions did not differ notably.

To obtain a concise but principled characterization of the developmental trajectory, we additionally performed a Bayesian change point analysis, using the package `RBeast` [@zhao2019detecting]. We sought the most likely change points in our data, assuming a constant mean (i.e., a flat line, zero degree polynomial) within each segment. To avoid "overreactions" to outlying data points, we constrained the model to have minimally 10 data points between consecutive change points (= half of the data points collected per adult decade). We inspected the posterior probability of different numbers of change points and the locations of these change points (mean and 95% CrI).[^2]

[^2]: In a supplementary analysis, we varied the parameters of our change point analysis and modified the number of allowed change points, the minimum number of data points between change points, and the polynomial order. When we allowed more explorative room, the models became more sensitive and added more fine-grained change points. The exact location of the change points varied slightly. However, the overall interpretation stayed the same, fitting our initial visual inspection: while early childhood was characterized by much change, adults showed a relatively stable level of imprecision, with a minor decay toward elderly adulthood. See Supplements for further detail.

@@

(ref:figlab1) **Developmental trajectory of gaze understanding across the lifespan**. Grey dots show the mean level of imprecision (ie., absolute distance between the target's center and the participant's click) for each subject averaged across trials. The unit of imprecision is counted in target width, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Blue lines show 100 draws from the expectation of the posterior predictive distribution of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the locations of the most prominent changes according to our Bayesian change point analysis.

@@

High levels of variation pointed to substantial individual differences in all age groups (overall imprecision mean = `r lifespan_variation$imprecision_mean`, sd = `r lifespan_variation$imprecision_sd`, range = [`r lifespan_variation$imprecision_min` - `r lifespan_variation$imprecision_max`]). We found substantial evidence for a non-linear development in gaze understanding across the lifespan. The Gaussian process model was clearly preferred over the polynomial models due to the highest predictive accuracy according to the LOO ELPD estimates: elpd_diff between Gaussian process and cubic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mcubic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mcubic"]`); elpd_diff between Gaussian process and quadratic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mquadratic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mquadratic"]`); elpd_diff between Gaussian process and linear model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mlinear"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mlinear"]`); all in favor of the Gaussian process model. Moreover, the Gaussian Process model showed the greatest model weight (approximating `r lifespan_mcomparison$modelweight[lifespan_mcomparison$model == "mgaussianprocess"]`). For the imprecision in gaze understanding, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) indicated nonlinearity.

The Bayesian change point analysis revealed `r lifespan_beast$trend$ncp_mode` major shifts in gaze understanding during the lifespan (MAP estimate with `r lifespan_beast$trend$ncpPr[lifespan_beast$trend$ncp_mode + 1] * 100 %>% round(2)`% probability). The change points occurred at `r lifespan_cps$cp[1]` (95% CrI [`r lifespan_cps$lowerCrI[1]`; `r lifespan_cps$upperCrI[1]`]); `r lifespan_cps$cp[2]` (95% CrI [`r lifespan_cps$lowerCrI[2]`; `r lifespan_cps$upperCrI[2]`]); `r lifespan_cps$cp[3]` (95% CrI [`r lifespan_cps$lowerCrI[3]`; `r lifespan_cps$upperCrI[3]`]); `r lifespan_cps$cp[4]` (95% CrI [`r lifespan_cps$lowerCrI[4]`; `r lifespan_cps$upperCrI[4]`]); `r lifespan_cps$cp[5]` (95% CrI [`r lifespan_cps$lowerCrI[5]`; `r lifespan_cps$upperCrI[5]`]); and `r lifespan_cps$cp[6]` years (95% CrI [`r lifespan_cps$lowerCrI[6]`; `r lifespan_cps$upperCrI[6]`]). In short: we found a rapid initial improvement in gaze understanding in early childhood, followed by a long period of minor, very slow change with slightly increasing levels of imprecision toward old age.

@@

We investigated the shape of change in gaze understanding across the lifespan and found a non-linear developmental trajectory, in which young children quickly enhanced their level of proficiency. Performance peaked around early adulthood, while there was a minor decay in later adulthood. This is consistent with the view that humans fine-tune their existing gaze understanding ability after the first emergence in infancy. Furthermore, we observed substantial individual differences in all age groups. While variation was highest in the three- and four-year-olds, it remained relatively stable across the lifespan.

>>>>

---
title: "Variation in gaze following across the life span: A process-level perspective"
shorttitle: "modeling variation in gaze following"
author:

  - name          : "Julia Christin Prein"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Max Planck Institute for Evolutionary Anthropology, Deutscher Platz 6, 04103 Leipzig, Germany"
    email         : "julia_prein@eva.mpg.de"
    role          :
      - Conceptualization
      - Methodology
      - Software
      - Investigation
      - Formal Analysis
      - Writing - Original Draft Preparation
      - Writing - Review & Editing

@@

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze following across the lifespan (*N* = 471 3- to 80-year-olds; children from a mid-sized German city; international remotely tested adults). We found a steep performance improvement during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze following in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we empirically tested how gaze following relates to vector following in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze following is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze following, individual differences, cognitive modeling, lifespan"
wordcount         : "X"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
classoption       : "man"

header-includes:
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
  - |
    \makeatletter
    \renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-1em}%
      {\normalfont\normalsize\bfseries\typesectitle}}

    \renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
      {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
      {-\z@\relax}%
      {\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
    \makeatother

csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            : papaja::apa6_word

editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

# Variation in gaze following across the life span: A process-level perspective

@@

**Data availability statement:** The gaze following task (<https://ccp-odc.eva.mpg.de/tango-demo/>) is open source (<https://github.com/ccp-eva/tango-demo>). The data sets generated during and/or analyzed during the current study are available in the following repository (<https://github.com/ccp-eva/gazecues-modeling>). All experiments and analyses were pre-registered prior to data collection (<https://osf.io/zjhsc/>).

@@

**Funding:** This study was funded by the Max Planck Society for the Advancement of Science, a noncommercial, publicly financed scientific organization (no grant number). Manuel Bohn was supported by a Jacobs Foundation Research Fellowship (grant no. 2022-1484-00).

@@

-	Gaze following develops beyond infancy. Highest precision levels in localizing attentional foci are reached in young adulthood with a slight decrease towards old age. 
-	We present a computational model that describes gaze following as a process of estimating pupil angles and the corresponding gaze vectors. 
- The model explains individual differences and recovers signature patterns in the data. To estimate the relation between gaze- and vector following, we designed a non-social vector following task. 
-	We found substantial correlations between gaze following and vector following, as well as Level 2 perspective-taking. Other Theory of Mind tasks did not correlate.

# Introduction
Following the gaze of others is valuable for extracting information from the environment. It guides us to "informational hotspots" [@meltzoff2010social, p.1] and can be used to identify internal states such as intentions or emotions [@corkum1998origins; @pfeiffer2013gaze]. As one of the most fundamental social-cognitive abilities, gaze-following is an integral part of almost every form of social interaction, including communication, collaboration, cultural and social learning [@emery2000eyes; @rakoczy2022foundations; @frith2012mechanisms; @tomasello2003makes; @bohn2019pervasive; @moore2008development; @hessels2020how], and has been extensively studied in infancy [for review, see @delbianco2019developmental]. In the most commonly used paradigm [e.g., @astor2020social; @gredeback2010development; @byers-heinlein2021development; @ishikawa2022physiological], the experimenter looks directly at the infant before shifting their head and eyes to one of two objects. Infants’ looking times to the target or the proportion of choosing the target over the distractor are measured. Research in this tradition finds that infants as young as three to four months can follow the gaze of another agent [@delbianco2019developmental; @dentremont2000perceptual; @dentremont1997demonstration; @astor2021gaze].

Previous research has shown a refinement of gaze-following abilities in a child's first and second year of life [e.g., @astor2021gaze; @butterworth1991minds; @brooks2002importance]. At the end of their first year of life, infants can follow gaze to locations outside their current visual field and move themselves to gain proper perceptual access [@moll200412; @deak2000effects; @butterworth1991minds; @corkum1995development]. However, we do not know much about the developmental progression beyond these qualitative milestones. One possibility is that the ability to follow gaze does not improve beyond infancy. Yet, most, if not all, cognitive abilities continue to develop throughout childhood [e.g., @gathercole2004structure; @gredeback2010development]. It seems likely that children fine-tune their gaze following as they get older - presumably while using them in social interactions. To capture the development in gaze following beyond infancy, Study 1 was included participants from preschool to old age.

Up until today, only a handful of gaze following studies differentiate between manipulating head and eye movement. @michel2021effects found that gaze following in four-month-olds was likely driven by other's head- instead of eye movements. @corkum1995development, @lempers1979young, and @lempers1977development suggest that infants, at least until 19 months, struggle when eye and head direction diverge. From farther distance, body or face orientation can act as more salient cues to determine another's area of attention. However, eye direction indicates a more precise location of focus [@emery2000eyes; @stiefelhagen2002head; however, see @loomis2008psychophysics for peripheral vision], and allows to anticipate likely future actions [@zohary2022gaze; @friesen2011gaze]. The three studies included in this work, therefore, focused on subtle gaze cues and isolated eye movement alone.

Group-level analyses of gaze-following abilities (e.g., average age at which children as a group reach an above-chance performance) may mask individual differences between children. Measuring individual differences in basic aspects of social cognition is important to understand the underlying processes and to quantify the impact of environmental influences and other cognitive abilities [@birch2017perspectives; @delbianco2019developmental]. Across the three studies, we measured gaze following continuously by using a task that is designed to capture individual-level variation [@prein2023tango]: the TANGO avoids floor and ceiling effects in children and adults and is thus particularly suited to examine how gaze following changes with age.

A promising approach to interpreting individual differences in cognitive abilities is computational cognitive modeling. Existing computational models have described gaze following via reinforcement learning [@ishikawa2020learning], as a consequence of goal interference and mapping self-experience onto other agents [@friesen2011gaze] or an interplay of object distances/saliencies and head poses [@lau2004learning; @jasso2006using; @recasens2015they]. As such, these models focus on the motivation behind gaze following or on situations in which the context (e.g., head orientation, surrounding objects) offers cues to the gaze direction. Our goal, however, was to formulate a theory that models how people estimate gaze direction based on the eyes alone (i.e., “literal” gaze following) when no target objects are present [^1] and the other’s head is frontally oriented.

[^1]: As @jasso2006using put it: “There is a general agreement that tests of gaze following in the absence of targets are more stringent than with targets, because that eliminates the possibility that infants are simply following the target’s saliency” (p. 2).

To our knowledge, there are three views that focus on the eyes alone that conceptualize gaze as (1) a beam, (2) a cone, or (3) a line (alternatively called a vector, ray, or line-of-sight). @guterstam2019implicit proposed the idea of gaze as a force-carrying beam. This theory, however, focuses on how people's implicit assumptions about the physical properties of an object changes when someone looks at it. The idea of gaze as a cone is mostly concerned with the question of how people determine whether someone else looks at them [@horstmann2021examining; @gamer2007you]. The conception most relevant to the present study sees eye gaze as a line: @yaniv1990heuristics proposed that children extrapolate an imaginary trajectory between the agent and the object to identify the focus of attention (similar to “geometrical” gaze following; [@butterworth1991minds]). Anecdotal evidence was already reported by @walker1977perspective, who observed two children pointing their fingers into the air, drawing a line between an agent and an object, and saying, “He sees that” (p. 354). @michelon2006two found that response time in Level 1 visual perspective-taking tasks depends on the distance between the agent and the object (i.e., the length of the line-of-sight). Some researchers additionally highlight iris eccentricity, that is, the ratio of the visible sclera on each side of the pupil [@anstis1969perception; @todorovic2006geometrical; @symons2004you]. @todorovic2006geometrical defined gaze direction as “the vector positioned along the visual axis, pointing from the fovea of the looker through the center of the pupil to the gazed-at spot” (p. 3550). @symons2004you furthermore reasoned that “the perceiver must use the asymmetrical configuration of the dark-white contrast of another individual’s eyes, and trace along two invisible sight-lines to their convergent point, that is, the third part of the triad (e.g., an object or a person)” (p. 452). Based on their finding that adults’ gaze direction sensitivity decreases when only one eye is shown, @symons2004you conclude that information from both eyes must be integrated.

While the previously mentioned conceptualizations focus on the direction of eye gaze, they (A) cannot explain how people differ in their abilities to precisely estimate gaze direction, and (B) are not clearly expressed as formal, mathematical models with explicit assumptions and testable predictions. In the words of [@gamer2007you]: “Given the social relevance of determining gaze direction, the psychophysics of gaze is underdeveloped” (p. 705).

Here, we propose a cognitive model of gaze following, which builds upon the notion of eye gaze as line-of-sight tracing and extends this by explicitly modeling individual differences. Our gaze model assumes participants infer the locus of someone’s attention to be where two estimated gaze vectors meet. Each of these gaze vectors results from connecting the center of the agent’s eyeball and the center of the pupil. Because the center of the eyeball is not directly observable, vector estimation happens with a degree of uncertainty. Development of gaze following corresponds to a decrease in uncertainty. Individual differences correspond to systematic differences in uncertainty.

By focusing on individual differences, we can further address the relationship between gaze following and other cognitive abilities. A longstanding question has been whether gaze following is related to Theory of Mind (ToM) [e.g., @brooks2015connecting]. @moll2011perspectivetaking have suggested that joint attention (including gaze following) might be seen as “Level 0 perspective-taking”, which provides the foundation for later-emerging, more complex perspective-taking abilities. On the other hand, an alignment of infants’ visual attention to another’s gaze does not necessarily indicate understanding the intentions of the agent [@aslin2007look]. Infants could simply align their orientation without processing what exactly the other is seeing [@butterworth1991minds]. In fact, one might question if such an alignment reflects an understanding of visual perspectives at all because the “target” or “object of representation” is not necessarily specified [@perner2003perspective, p. 358]. Consequently, @astor2022gaze have listed the relationship between gaze following and perspective-taking as one of their five big open questions in gaze following research. Therefore, in Study 3, we assessed how gaze following relates to ToM abilities, especially visual perspective-taking.

Taken together, the present study had three main goals: first, we studied the development of gaze following beyond infancy (Study 1). Instead of capturing the youngest age at which children follow gaze, we examined how this ability changes with age. Our second goal was to provide a process-level theory of gaze following – and, most importantly, individual differences therein. We proposed a computational cognitive model, which formalized gaze following as a form of vector following, and tested whether our model explained empirical data (Study 2). Third, we examined which (social-)cognitive components comprise gaze following (Study 3). Based on our model, we predicted that gaze following should be related to non-social vector following. Additionally, we assessed the link between gaze following and ToM measures, with a particular focus on visual perspective-taking.

# Study 1: Gaze following across the lifespan

The study was pre-registered prior data collection: <https://osf.io/snju6> (child sample) and <https://osf.io/6yjz3> (adult sample). The study obtained ethical clearance by the MPG Ethics commission in Munich, Germany, falling under an umbrella ethics application (Appl. No. 2021_45). Data was collected between May 2021 and April 2023.

@@

We collected data online from 3- to 80- year-olds (see Supplements for further details). The child sample consisted of `r sum(lifespan_sample$n_total[lifespan_sample$age_group < 18])` participants and was recruited via an internal database of families in Leipzig, Germany, who volunteered to participate in child development studies. Participants came from ethnically homogeneous, mixed socioeconomic backgrounds with mid to high parental education levels. They lived in an industrialized, urban Central-European context in a mid-size German city (approx. 600,000 inhabitants; median individual monthly net income approx. 1,600€ as of 2021). Most were raised monolingually in a nuclear two-generational family setting. Information on demographics and socioeconomic status was not formally recorded on a participant level.

@@

We used the continuous version of the TANGO [@prein2023tango]. The task was presented as a web application (demo [https://ccp-odc.eva.mpg.de/tango-demo/](https://ccp-odc.eva.mpg.de/tango-demo/.); source code <https://github.com/ccp-eva/tango-demo>). The TANGO showed satisfactory internal consistency and retest reliability (*Pearson's r* from .7 to .8; @prein2023tango) and no floor or ceiling effects for children and adults.

@@

We fit GLMMs that make different assumptions about the developmental trajectory, modeling the relationship between age and performance as linear, quadratic, or cubic. In addition, we fit a Gaussian Process model [@burkner2017brms], which assumes a smooth relationship but avoids enforcing a particular shape. Per individual, imprecision was aggregated across trials and modeled as a `lognormal` distribution.[^2] The unit of imprecision was counted in target widths, i.e., an imprecision of 1 meant clicking one balloon width to the left or right of the true target center. We inspected the posterior distributions (mean and 95% Credible Interval (CrI)) for the age estimates and compared models via model weights and the difference in expected log pointwise predictive density (ELPD) estimated using leave-one-out cross-validation (LOO) [@vehtari2017practicala].

[^2]: Originally, we fit our models on a trial-by-trial basis with the following structure: `performance ~ age + symmetricPosition + trialNr + (1 + symmetricPosition + trialNr | subjID)`. However, the Gaussian Process model was computationally heavy. Thus, we simplified the model structure, aggregated data on a subject level, and included only age as an effect. See Supplements for a comparison between the original and the here-reported model structures. The model predictions did not differ notably.

To obtain a concise but principled characterization of the developmental trajectory, we additionally performed a Bayesian change point analysis, using the package `RBeast` [@zhao2019detecting]. We sought the most likely change points in our data, assuming a constant mean (i.e., a flat line, zero-degree polynomial) within each segment. To avoid "overreactions" to outlying data points, we constrained the model to have minimally 10 data points between consecutive change points (= half of the data points collected per adult decade). We inspected the posterior probability of different numbers of change points and the locations of these change points (mean and 95% CrI).[^3]

[^3]: In a supplementary analysis, we varied the parameters of our change point analysis and modified the number of allowed change points, the minimum number of data points between change points, and the polynomial order. When we allowed more explorative room, the models became more sensitive and added more fine-grained change points. The exact location of the change points varied slightly. However, the overall interpretation stayed the same, fitting our initial visual inspection: while early childhood was characterized by much change, adults showed a relatively stable level of imprecision, with a minor decay toward elderly adulthood. See Supplements for further detail.

@@

(ref:figlab1) **Developmental trajectory of gaze following across the lifespan**. Grey dots show the mean level of imprecision (ie., absolute distance between the target's center and the participant's click) for each subject averaged across trials. The unit of imprecision is counted in target width, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Blue lines show 100 draws from the expectation of the posterior predictive distribution of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the locations of the most prominent changes according to our Bayesian change point analysis.

@@

High levels of variation pointed to substantial individual differences in all age groups (overall imprecision mean = `r lifespan_variation$imprecision_mean`, sd = `r lifespan_variation$imprecision_sd`, range = [`r lifespan_variation$imprecision_min` - `r lifespan_variation$imprecision_max`]). We found substantial evidence for a non-linear development in gaze following across the lifespan. The Gaussian process model was clearly preferred over the polynomial models due to the highest predictive accuracy according to the LOO ELPD estimates: elpd_diff between Gaussian process and cubic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mcubic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mcubic"]`); elpd_diff between Gaussian process and quadratic model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mquadratic"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mquadratic"]`); elpd_diff between Gaussian process and linear model = `r lifespan_mcomparison$elpd_diff[lifespan_mcomparison$model == "mlinear"]` (SE = `r lifespan_mcomparison$se_diff[lifespan_mcomparison$model == "mlinear"]`); all in favor of the Gaussian process model. Moreover, the Gaussian Process model showed the greatest model weight (approximating `r lifespan_mcomparison$modelweight[lifespan_mcomparison$model == "mgaussianprocess"]`). For the imprecision in gaze following, the standard deviation (*b* = `r lifespan_gp$estimate[lifespan_gp$param == "sdgp(gpage_centered)"]`, 95% CrI [`r lifespan_gp$l_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`; `r lifespan_gp$u_95_percent_ci[lifespan_gp$param == "sdgp(gpage_centered)"]`]) indicated nonlinearity.

The Bayesian change point analysis revealed `r lifespan_beast$trend$ncp_mode` major shifts in gaze following during the lifespan (MAP estimate with `r lifespan_beast$trend$ncpPr[lifespan_beast$trend$ncp_mode + 1] * 100 %>% round(2)`% probability). The change points occurred at `r lifespan_cps$cp[1]` (95% CrI [`r lifespan_cps$lowerCrI[1]`; `r lifespan_cps$upperCrI[1]`]); `r lifespan_cps$cp[2]` (95% CrI [`r lifespan_cps$lowerCrI[2]`; `r lifespan_cps$upperCrI[2]`]); `r lifespan_cps$cp[3]` (95% CrI [`r lifespan_cps$lowerCrI[3]`; `r lifespan_cps$upperCrI[3]`]); `r lifespan_cps$cp[4]` (95% CrI [`r lifespan_cps$lowerCrI[4]`; `r lifespan_cps$upperCrI[4]`]); `r lifespan_cps$cp[5]` (95% CrI [`r lifespan_cps$lowerCrI[5]`; `r lifespan_cps$upperCrI[5]`]); and `r lifespan_cps$cp[6]` years (95% CrI [`r lifespan_cps$lowerCrI[6]`; `r lifespan_cps$upperCrI[6]`]). In short: we found a rapid initial improvement in gaze following in early childhood, followed by a long period of minor, very slow change with slightly increasing levels of imprecision toward old age.

@@

We investigated the shape of change in gaze following across the lifespan and found a non-linear developmental trajectory, in which young children quickly enhanced their level of proficiency. Performance peaked around early adulthood, while there was a minor decay in later adulthood. These results support the idea that humans fine-tune their existing gaze following ability after the first emergence in infancy. Furthermore, we observed substantial individual differences in all age groups. While variation was highest in the three- and four-year-olds, it remained relatively stable across the lifespan.
